Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for Vietnamese and International Students

### 2023-06-12 phd-cs-us.tex
updates

However, I understand that many international students do not have the opportunities to publish in top places, so general confs/journals would suffice.  But be sure to upload your papers with your application and talk about it in your \hyperref[sec:research-statement]{statement}.

@@

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements},  so I will not talk too much about it. In short, discuss about your research vision and convince us that you can achieve it through your experience, e.g., published papers, or if you work on some projects by yourself, talk about it.

@@

Even if you have other funding and do not need TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for research career where you often have to give talks and tell people about your work. Note that GMU sometimes has classes that a more senior student can teach.  In that case, you will be paid as a lecturer, which is higher than GTA.  This is a good opportunity for students to get teaching experience and also get paid more.
\subsection{Research Assistant (RA)}
RA is provided through a professor through their own funding so you can work on their project.  
You do not need to teach as an RA, so you can focus on your research. Depending on the professor, RA may be available during the summer.

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school X that you don't know much about has very strong research in your interested field Y. This is also a good way to learn about individual faculty (who works on what) and the well-known CS conferences\footnote{In CS (and only in CS), conferences, not journals, are often the main venue to publish research finding.}. \S\ref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.
\begin{commentbox}
    Dat: MOst Vietnamese students, including those from top schools, do not know about CSRankings.  May be applicants who worked at top research places such as VinAI would know about it.
\end{commentbox}

@@

\section{How much do \emph{you} cost?}
PhD students often wonder how their salary is so low compared to their advisor's ludicrous grants or that why their offer letters sometime mentioned their benefits worth way more than what they actually receive (i.e., stipend).  This section aims to shed some light to these questions by breaking down how much a GRA costs to their advisor.

Tab.~\ref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US.  Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$27000 (GMU actually pays more) and therefore a 3-month summer of \$9000. I will also use GMU tuition rate of about \$15000/year for full-time study and a 58.9\% \emph{indirect cost}, which is what GMU charges for administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student will make 2 conference trips per year, one domestic and one international (so conf. registration, airline tickets, taxi, meals, etc are all included). At the end, the total budget comes out to be \$69K.  \emph{The summary is that while you're paid X, your advisor probably pays 2X for you}.

\begin{table}
    \centering
    \small
    \caption{GRA cost breakdown. F \& A is Facilities \& Administrative Cost Base and 
    MTDC is Modified Total Direct Cost.}\label{tab:cost}
    \begin{tabular}{rcl}
        \textbf{Budget} & \textbf{Cost} & \textbf{Notes} \\
        \midrule
        GRA (9-month) & 27K & \\
        GRA (summer)  &9K	& 3-month, 20hrs/week\\
        \textbf{Total Salary} &\red{36K}	&\\
        \midrule
        Health Insurance	&3K	& full year\\
        Tuition (In-State) &	15K	& (\$680/ Credit + \$150/Student Fee/ Credit)* 9 credits = \\
        &&\$7470 (\$6120 + \$1350) per semester\\
        \textbf{Total Tuition \& Insurance}	&\red{18K}	&Full year tuition + insurance\\
        \midrule
        Conference Registration	& 500 & \\
        International Travel &	1800& \\
        Domestic Travel	& 700	& \\
        \textbf{Total Travel}&	\red{3K}	\\
        \midrule
     Total Direct Cost	& \red{57K}	&Salary + Travel + Health + Tuition \\
     F \& A (MTDC)	& 21K	& Direct Cost - GRA Salary\\
    Total Indirect Cost	& \red{12K}	&58.9\% of MTDC\\
    \textbf{Total (Direct + Indirect)} &	\red{69K}	& Budget for a GRA\\
        \bottomrule
    \end{tabular}
\end{table}

>>>>

However, I understand that many international students do not have the opportunities to publish in top places, so general confs/journals would suffice.  But be sure to upload your papers with your application and talk about them in your \hyperref[sec:research-statement]{statement}.

@@

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, discuss about your research vision and convince us that you can achieve it through your experience, e.g., published papers, or if you work on some projects by yourself, talk about it.

@@

Even if you have other funding and do not need TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for research career where you often have to give talks and tell people about your work. Note that GMU sometimes has classes that a more senior student can teach.  In that case, you will be paid as a lecturer, which is higher than GTA.  This is a good opportunity for students to get teaching experience and also get paid more.
\subsection{Research Assistant (RA)}
RA is provided through a professor through their own funding so you can work on their project.  
You do not need to teach as an RA, so you can focus on your research. Depending on the professor, RA may be available during the summer. \S\ref{sec:ra-cost} gives more details on RA budget.

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school X that you don't know much about has very strong research in your interested field Y. This is also a good way to learn about individual faculty (who works on what) and the well-known CS conferences\footnote{In CS (and only in CS), conferences, not journals, are often the main venue to publish research finding.}. \S\ref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.
\begin{commentbox}
    Dat: Most Vietnamese students, including those from top schools, do not know about CSRankings.  May be applicants who worked at top research places such as VinAI would know about it.
\end{commentbox}

@@

\section{How much do \emph{you} cost?}\label{sec:ra-cost}
PhD students often wonder how their salary is so low compared to ludicrous grants their advisors get or that why their offer letters sometime mentioned that their benefits worth way more than what they actually receive (i.e., stipend).  This section aims to shed some light to these questions.

Tab.~\ref{tab:cost} shows the budget breakdown for a GRA per year (this level of details is what faculty actually uses when applying for funding).
These numbers are based on my experience at public universities in the US.  Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$27000 (GMU actually pays more) and therefore a 3-month summer of \$9000. I will also use GMU tuition rate of about \$15000/year for full-time study and a 58.9\% \emph{indirect cost}, which is what GMU charges for administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student will make 2 conference trips per year, one domestic and one international (so conf. registration, airline tickets, taxi, meals, etc are all included). At the end, the total budget comes out to be \$69K.  \emph{The summary is that while you're paid X, your advisor probably pays 2X for you}.

\begin{table}
    \centering
    \small
    \caption{GRA cost breakdown. F \& A is Facilities \& Administrative Cost Base and 
    MTDC is Modified Total Direct Cost. These are things that the university can charge overhead to.}\label{tab:cost}
    \begin{tabular}{rcl}
        \textbf{Budget} & \textbf{Cost \$} & \textbf{Notes} \\
        \midrule
        GRA (9-month) & 27K & \\
        GRA (summer)  &9K	& 3-month, 20hrs/week\\
        \textbf{Total Salary} &\red{36K}	&\\
        \midrule
        Health Insurance	&3K	& full year\\
        Tuition (In-State) &	15K	& (\$680/ Credit + \$150/Student Fee/ Credit)* 9 credits = \\
        &&\$7470 (\$6120 + \$1350) per semester\\
        \textbf{Total Tuition \& Insurance}	&\red{18K}	&Full year tuition + insurance\\
        \midrule
        Conference Registration	& 500 & \\
        International Travel &	1800& \\
        Domestic Travel	& 700	& \\
        \textbf{Total Travel}&	\red{3K}	\\
        \midrule
     Total Direct Cost	& \red{57K}	&Salary + Travel + Health + Tuition \\
     F \& A (MTDC)	& 21K	& Direct Cost - GRA Salary\\
    Total Indirect Cost	& \red{12K}	&58.9\% of MTDC\\
    \textbf{Total (Direct + Indirect)} &	\red{69K}	& Budget for a GRA\\
        \bottomrule
    \end{tabular}
\end{table}
